A Beginners Guide to Perfomance Factors of Machine Learning Learning Applications with Focus on Neural Nets and Deep Learning

### 2018-01-06 paper.tex
add Choice of Neural Network section and further citing in background(still not final though)

@@

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{
%Performance Factors for Large Scale Machine Learning Applications
A Beginners Guide to Perfomance Factors of Machine Learning Learning Applications with Focus on Neural Nets and Deep Learning
}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
%\begin{itemize}
%\item Deliver short introduction for neural networks with different aspect focused examples
%\item What we want to show in this paper(performance factors)
%\item To what end we came
%\item some other stuff that comes up during the paper and is important enough to be named at the beginning
%\end{itemize}

@@

A more advanced architecture is to increase the number of hidden layers to enlarge the application possibility of these networks for more complex models these are then called Deep Networks. The most commonly used ones are Variational Auto Encoders , Deep Belief Networks, Generative Adversial Networks, Recurrent Neural Networks ,e.g. Long short-term memory networks, Deep Convolutional Neural Networks, Deconvolutional Networks.

@@

The most common neural network frameworks which we, because of their wide popularity, will mostly view are Tensorflow\cite{abadi2016tensorflow}, Caffe\cite{jia2014caffe}, Torch\cite{collobert2002torch}, CNTK, deeplearning4j, Keras, MXNET. These frameworks have been benchmarked in numerous settings and applications and can help us show performance influencing factors.

@@

@@

\section{Choice of Neural Nets}	
what net when to use, and starting values of neural nets like amount of layers and starting weights

\section{Choice of Processing Unit}
The next important choice is the choice of the actual operating calculation unit. We distinguish them into three main models, CPU, GPU and GPU-Clustering.
\subsection{CPU}
As the first most naive option we looked on the CPU for training of the neural net with the simple single thread and more complex multi-thread calculation.
\subsubsection{Single-thread}
The most naive implementation is the single-thread CPU calculation, because oft this more  brute force approach and sequential execution and calculation this is also the slowest and with a larger number of connected nodes simply unfeasible because of the excessive calculation time required . But as can be thought of this does not uitilize the full capacity of the CPU and therefore results in performance loss in comparison to multithread usage(Fig. 1).
\subsubsection{Multi-thread}
A more complex approach consist of utilizing the multi-core characterisitics of modern CPUs, by splitting of training calculation with only requirements to already calculated values. This is for example preferrable when training with the help of a \emph{Backpropagation} algorithm. The effect  continues to grow by using more threads but reaches it limits when using more threads than physical cores available.
\begin{figure} 
\centering
\begin{tabular}{c c c c c}
\hline
Framework & 1 Thread & 2 Threads &4 Threads & 8 Threads\\\hline
Caffe & 1.324 & 0.790 & 0.578 & 15.444 \\
Tensorflow & 7.062 & 4.789 & 2.648 & 1.938 \\
Torch & 1.329 & 0.710 & 0.423 & na \\\hline
\end{tabular}
\caption{Training time(in s) for a mini-batch of size 64 of a Fully Connected Network(FCN) trained with synthetic data on a i7-3820 with 4-physical cores using 3 different commonly used Deep Learning frameworks\cite{shi2016benchmarking}}
\label{fig_ttfcn}
\end{figure}

Though this approach is limited by the actual core number available and should not overstep the number of physical cores, like with intel \emph{hyper-threading} technology, which can lengthen the required calculation time because of a more inefficient usage\cite{shi2016benchmarking}.
\subsection{GPU}
A more advanced implementation is to use the shared memory, multi-core environment of modern GPUs. For example with NVIDIAs CUDA this can be done efficiently and result in a speed-up of calculation of up to 3 times the CPU-based approachs performance for deep learning algorithms over optimized floting-point baseline\cite{shi2016benchmarking}. Allmost all recent neural net learning frameworks support this calculation unit because of its excellent performance rating for neural networks.

@@

%\begin{figure}
%\centering
%\includegraphics[width=0.2\textwidth]{a.png}
%
%\caption{Time per mini-batch in s compared between different amount of GPUs used\cite{shi2016benchmarking}}
%\label{fig_m_gpu}
%\end{figure}

@@

\subsubsection{Multiple GPU-Clusters}
Taking a step further it is possible to connect multiple Clusters of GPUs to compute a single task. In doing this we again increase raw computation power, but also increase the total amount of communication required.

@@

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively. 
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

>>>>

%%%%IMAGE PACKAGE
\usepackage{graphicx}

@@

\begin{document}

\title{
%Performance Factors for Large Scale Machine Learning Applications
A Beginners Guide to Perfomance Factors of Machine Learning Learning Applications with Focus on Neural Nets and Deep Learning
}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}

@@

A more advanced architecture is to increase the number of hidden layers to enlarge the application possibility of these networks for more complex models these are then called Deep Networks. The most commonly used ones are Variational Auto Encoders , Deep Belief Networks, Generative Adversial Networks, Recurrent Neural Networks ,e.g. Long short-term memory networks, Deep Convolutional Neural Networks, Deconvolutional Networks, and Recursive Neural Networks.

@@

The most common neural network frameworks which we, because of their wide popularity, will mostly view are Tensorflow\cite{abadi2016tensorflow}, Caffe\cite{jia2014caffe}, Torch\cite{collobert2002torch}, CNTK\cite{gitcntk}, deeplearning4j\cite{websitedl4j}, Keras\cite{websitekeras}, MXNET\cite{websitemxnet}. These frameworks have been benchmarked in numerous settings and applications and can help us show performance influencing factors.

@@

%%%%BEGIN PROCESSING UNIT

TODO

%%%%END PROCESSIONG UNIT

@@

\section{Choice of Neural Network}

%%%%BEGIN IMPORTANCE
The first important choice is to what neural network is going to be used. This decision is heaviliy reliant on the application that is to be implemented. This decision influences all further actions and has to be carefully made, to assist this decision we will offer some generalization when to use simple neural networks or deep neural network and what specializations of certain neural networks are giving them advantage while doing specific tasks.
%%%%END IMPORTANCE

%%%%BEGIN ADVANTAGES DISADVANTAGES SIMPLE AND DEEP
\subsection{What kind of networks to use}
\subsubsection{Simple Networks}
Simple task can be done by a network which is not deep like a FFN or a RBM. They are faster to train and even larger training sets can be completed in a reasonable time without multithreading or usage of GPUs. But there small size can lead that accuracy decreses when the model becomes more difficult than anticipated. Small scale applications often incur that a simple neural network is used because of the smaller training set and shorter computation time.

\subsubsection{Deep Networks}
More complex tasks more often require a deep neural network that can learn a more complex model than simple networks and offer a greater accuracy while doing this. The major downside of these networks that they require a larger amount of time to train and can be oversized for the actual to be represented model, and therefore adding unnecessary computation time. Large scale applications mostly contain deep networks because of their larger training sets and more complex models underlying the actual task.

We can not offer a clear answer when to use which kind of network, every problem must be individually analyzed and based on the information, and complexity, as well as the acceptable error rate of the final network the decision has to be made. For cases that are unclear a simple network can first be constructed,because of their short training time, and if the result is unsatisfying a more complex network can be considered.
%%%%END ADVANTAGES DISADVANTAGES SIMPLE AND DEEP

%%%%INTITAL VALUES AND AMOUNT OF NODES

An also important choice is the setting of the initial values of connection weights, here it is advisable to start at low values like $0.1$ to prevent straying too far from the optimal solution, since algorithm like backpropagation mostly find a local optimum initial values too far of from the global optimum will deliver worse results.

%%%%END INITIAL VALUES AND AMOUNT OF NODES

%%%%DETAILS SIMPLE AKA SMALL
\subsection{Simple network Nodes}
If a simple network shall be used the next choice will be the amount of nodes in the hidden layer. This is of course dependent one foremost the number of inputs and outputs, the number of hidden nodes should deviate to strong from them. How strong this deviation really is is dependent on the chosen neural network, for example a feed forward network profits when the number is equal or larger than the number of inputs, while an auto encoder probably has fewer hidden nodes. For a simple network this decision can be if too far from the optimum less fatal than for a deep network, while a simple network may lose some accuracy or gain additional training time a deep network can experience these effects at a greater scale.

%%%%END DETAILS SIMPLE AKA SMALL

%%%%DETAILS DEEP AKA LARGE AND SMALL
\subsection{Deep networks Nodes}
Like in simple networks the amount of nodes to be used is an important value that determines the final ability of the network to fulfill its task in the required accuracy and speed. The acutal number of nodes chosen at the end is of course a bit more complex in deep networks for example networks like DCN require a special node configuration based on their basic idea(step by step decreasing of nodes per layer until desired compression reached) , or DN(opposite of DCN). So how the node configuration looks like is strongly dependent on the net chosen.

%%%%END DETAILS DEEP AKA LARGE AND SMALL

%%%%SPECIFIC NETS

\subsection{Specific deep networks}

%%%%END SPECIFIC NETS

\section{Choice of Processing Unit}
The next important choice is the choice of the actual operating calculation unit. We distinguish them into three main models, CPU, GPU and GPU-Clustering.
\subsection{CPU}
As the first most naive option we looked on the CPU for training of the neural net with the simple single thread and more complex multi-thread calculation.
\subsubsection{Single-thread}
The most naive implementation is the single-thread CPU calculation, because oft this more  brute force approach and sequential execution and calculation this is also the slowest and with a larger number of connected nodes simply unfeasible because of the excessive calculation time required . But as can be thought of this does not uitilize the full capacity of the CPU and therefore results in performance loss in comparison to multithread usage(Fig. 1).
\subsubsection{Multi-thread}
A more complex approach consist of utilizing the multi-core characterisitics of modern CPUs, by splitting of training calculation with only requirements to already calculated values. This is for example preferrable when training with the help of a \emph{Backpropagation} algorithm. The effect  continues to grow by using more threads but reaches it limits when using more threads than physical cores available.
\begin{figure} 
\centering
\begin{tabular}{c c c c c}
\hline
Framework & 1 Thread & 2 Threads &4 Threads & 8 Threads\\\hline
Caffe & 1.324 & 0.790 & 0.578 & 15.444 \\
Tensorflow & 7.062 & 4.789 & 2.648 & 1.938 \\
Torch & 1.329 & 0.710 & 0.423 & na \\\hline
\end{tabular}
\caption{Training time(in s) for a mini-batch of size 64 of a Fully Connected Network(FCN) trained with synthetic data on a i7-3820 with 4-physical cores using 3 different commonly used Deep Learning frameworks\cite{shi2016benchmarking}}
\label{fig_ttfcn}
\end{figure}
Though this approach is limited by the actual core number available and should not overstep the number of physical cores, like with intel \emph{hyper-threading} technology, which can lengthen the required calculation time because of a more inefficient usage\cite{shi2016benchmarking}.
\subsection{GPU}
A more advanced implementation is to use the shared memory, multi-core environment of modern GPUs. For example with NVIDIAs CUDA this can be done efficiently and result in a speed-up of calculation of up to 3 times the CPU-based approachs performance for deep learning algorithms over optimized floting-point baseline\cite{shi2016benchmarking}. Allmost all recent neural net learning frameworks support this calculation unit because of its excellent performance rating for neural networks.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{a.png}

\caption{Time per mini-batch in s compared between different amount of GPUs used\cite{shi2016benchmarking}}
\label{fig_m_gpu}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{gpu_cluster_perf_win.png}
\caption{Time in h of different asynchronous configurations until reaching destined error rate(loss)\cite{sastre2017scalability}}
\label{fig_cl_gpus}
\end{figure}

\subsubsection{Multiple GPU-Clusters}
Taking a step further it is possible to connect multiple Clusters of GPUs to compute a single task. In doing this we again increase raw computation power, but also increase the total amount of communication required. Similar to multiple GPUs this leads at the beginning with 2 or more Clusters to a stronger improvement than later on with 8 to 16 Clusters (Fig. 3)

@@
